Hypercoercions and a Framework for Equivalence of Cast Calculi

### 2019-10-24 paper.tex
read sec3 the first time, lots of things to fix
camalize thm names

\begin{figure}
  \[
  \begin{array}{lrcl}
  \stxrule{\ell}{Maybe $l$}{\epsilon \mid l}
  \end{array}
  \]

@@

\figref{fig:HC-D} defines the functions $id$, $seq$, and $cast$ that
construct \lazyD{} hypercoercions, towards satisifying the Cast ADT.
\figref{fig:HC-D} also defines some auxiliary functions.

The $seq$ operator is defined in terms of coercion composition, which
is the key to compressing coercions and maintaining space efficiency.
A composition operator for coercions typically requires the target
type of the first coercion to match the source type of the second.
However, it is useful to relax this restriction for the D blame
tracking strategy.
%
We shall need optional blame labels, written $\ell$ , that range over
$\epsilon$ or $l$. Then we write composition of hypercoercions as $c_1
\fatsemi^\ell c_2$.  When both $c_1$ and $c_2$ are $\hyperCoercionI$,
their composition is also $\hyperCoercionI$.  If the head of $c_2$ is
a projection, the result is $c_2$.  Otherwise, we need a label to
build the projection. Since the head of $c_2$ is $\epsilon$, its
source type must be a pretype. Thus we know $\ell$ must be a label and
we put it in the projection. When $c_1$ ends with a failure, the
composition is $c_1$ itself. In all the remaining cases, $c_1$ does
not end with a failure.  When $m_1$ and $m_2$ have different top
constructors ($m_1 \not\smile m_2$), the result is a failure coercion,
which needs a label. When $c_2$ starts with a projection, it is to
blame for casting a value to a shallowly inconsistent type. When $c_2$
starts with a no-op, $\ell$ must be a label. And we blame it for
composing shallowly inconsistent hypercoercions. The last two cases
compose $m_1$ and $m_2$ with the auxilliary function $m \fatsemi^\ell
m$, which assumes its inputs have the same top constructor. The
definition of $m \fatsemi^\ell m$ is a straightforward structural
recursion. Going back to the last two cases of $c_1 \fatsemi^\ell
c_2$. When $c_2$ starts with a projection, we do not know whether the
target type of $m_1$ matches the source of $m_2$, although we know
they are shallowly consistent. So we use the $l$ in the projection to
compose the middles $m_1 \fatsemi^l m_2$. When $c_2$ starts with a
no-op, we compose the middles using $\ell$.

@@

The proof of correctness for hypercoercions (in
Section~\ref{secc:framework:monoid-correct}) relies on the following
two basic properties of hypercoercions.

\begin{proposition}[\lazyD{} hypercoercions form a monoid]
  \label{thm:hc-monoid}
  For all $c : T_1 \Longrightarrow T_2$,
  $c_1 : T_1 \Longrightarrow T_2$,
  $c_2 : T_2 \Longrightarrow T_3$, and
  $c_3 : T_3 \Longrightarrow T_4$,
  \begin{enumerate}
    \item $seq(id(T_1),c) = c$,
    \item $seq(c,id(T_2)) = c$, and
    \item $seq(seq(c_1, c_2), c_3) = seq(c_1, seq(c_2, c_3))$.
  \end{enumerate}
\end{proposition}
\begin{proof}
  Part (1) and (2) are by induction on $c$. Part (3) is by induction
  on $c_2$ and case analysis on $c_1$ then $c_3$.
\end{proof}

\begin{proposition}[\lazyD{} Identity Casts]
  For all $T$ and $l$, $cast(T,l,T) = id(T) $
\end{proposition}
\begin{proof}
  By induction on T.
\end{proof}

@@

\begin{proposition}[\lazyUD{} hypercoercions form a monoid]
  For all $c : T_1 \Longrightarrow T_2$,
  $c_1 : T_1 \Longrightarrow T_2$,
  $c_2 : T_2 \Longrightarrow T_3$, and
  $c_3 : T_3 \Longrightarrow T_4$,
  \begin{enumerate}
    \item $seq(id(T_1),c) = c$,
    \item $seq(c,id(T_2)) = c$, and
    \item $seq(seq(c_1, c_2), c_3) = seq(c_1, seq(c_2, c_3))$.
  \end{enumerate}
\end{proposition}
\begin{proof}
  See the Agda proof at the following URL:

@@

\begin{proposition}[\lazyUD{} Identity Casts]
  For all $T$ and $l$, $cast(T,l,T) = id(T) $
\end{proposition}
\begin{proof}
  By induction on T.
\end{proof}

@@

\begin{proposition}[Equivalence of two \lazyD{} Cast ADTs]
  \label{thm:surely-lazyD-eqv}
  Assume $C_1$ and $C_2$ implements Lazy D Cast ADT,
  \[
  eval_{\mathcal{S}(C_1)} = eval_{\mathcal{S}(C_2)}.
  \]
\end{proposition}
\begin{proof}
  Without lost of generality, assume $\judgeSreduce{C_1}{e}{o}$, prove 
  $\judgeSreduce{C_2}{e}{o}$.\\
  By the definition of $eval_{\mathcal{S}(C)}$, we know 
\[
\judgeSreduceTrans{C_2}{\sOOinspect{e}{\emptyset}{wrap(\kOOmt)}}{\sOOhalt{o}}
\]
  By induction on this reduction sequence and Lemma~\ref{thm:CEKS-bisim}, we 
  have 
\[
\judgeSreduceTrans{C_2}{\sOOinspect{e}{\emptyset}{wrap(\kOOmt)}}{\sOOhalt{o}}
\]
  that is, $\judgeSreduce{C_2}{e}{o}$
\end{proof}

@@

\begin{proposition}[$L$ properties]\label{thm:L-LazyD-monoid}\ 
  \begin{enumerate}
  \item $L$ is an instance of the Lazy D ADT
  \item $L$ is a monoid
  \item $cast(T,l,T) = id(T)$
  \end{enumerate}
\end{proposition}
\begin{proof}
  Part (1) is straightforward because property (1) and (2) are trivial and 
  that $appS$ basically repeats property (3) - (11).
  Part (2) is true because in $L$ casts are list, $seq(c,c)$ is list 
  appending, and $id(T)$ is the empty list. Lists is know to be a monoid 
  where the binary operation is appending and the identity element is the 
  empty list.
  Part (3) follows immediately from the definition of $cast(T,l,T)$
\end{proof}

\begin{lemma}[$L$ is correct wrt. \ineffCEKD]
  \label{lem:L-correct}
  \[
  \text{\evalEqv{\effCEK{L}}{\ineffCEKD}}
  \]
\end{lemma}
\begin{proof}
  Immedialy from Correctness of \effCEK{C} \ref{thm:surely-monoidic-eval} and 
  L properties \ref{thm:L-LazyD-monoid}.
\end{proof}

@@

\begin{theorem}[\effCEK{C} is correct wrt. \ineffCEKD{}]
  \label{thm:surely-lazyD-correct}
  Suppose $C$ is a \lazyD{} Cast ADT.  If
  $\judgetype{\emptyset}{e}{T}$ and $o : T$, then
  \[
  \text{\evalEqv{\effCEK{C}}{\ineffCEKD}}
  \]
\end{theorem}
\begin{proof}
  If we have a Lazy D Cast ADT and it is a monoid, this 
  proof is immediately from Lemma~\ref{thm:surely-monoidic-eval} and 
  Proposition~\ref{thm:surely-lazyD-eqv}. One such representation is to 
  represent casts as a list of triples of type, label, and type, where 
  $id(T)$ is the empty list and $seq(c,c)$ is the list append. Another such 
  representation is the \lazyD\ hypercoercion!
\end{proof}

@@

\begin{lemma}[\lazyD{} hypercoercion is a \lazyD{} Cast ADT]
  \label{thm:hc-surely-lazyD}
\end{lemma}
\begin{proof} See the supplementary material. \end{proof}

\begin{theorem}[\lazyD{} hypercoercion Respect \ineffCEKD]
  If $\judgetype{\emptyset}{e}{T}$ and $o : T$
  \[
  \text{\evalEqv{\effCEK{H}}{\ineffCEKD}}
  \]
\end{theorem}
\begin{proof}
  Immediately from Theorem~\ref{thm:surely-lazyD-correct} and 
  Lemma~\ref{thm:hc-surely-lazyD}.
  Alternatively, from Lemma~\ref{thm:surely-monoidic-eval},
  Lemma~\ref{thm:hc-surely-lazyD}, and 
  Proposition~\ref{thm:hc-monoid}.
\end{proof}

>>>>

\begin{figure}
  \[
  \begin{array}{lrcl}
  \stxrule{\ell}{Optional labels}{\epsilon \mid l}
  \end{array}
  \]

@@

\figref{fig:HC-D} defines the functions $id$, $seq$, and $cast$ that
construct \lazyD{} hypercoercions, towards satisifying the Cast abstract data 
type defined in Section~\ref{sec:framework}.
\figref{fig:HC-D} also defines some auxiliary functions.

The $seq$ operator is defined in terms of coercion composition, which
is the key to compressing coercions and maintaining space efficiency.
A composition operator for coercions typically requires the target
type of the first coercion to match the source type of the second.
However, it is useful to relax this restriction for the D blame
tracking strategy.
%
We shall need optional blame labels, written $\ell$ , that range over
$\epsilon$ or $l$. The label is mandatory when the target
type of the first coercion does not match the source type of the second. 
Then we write composition of hypercoercions as $c_1
\fatsemi^\ell c_2$.  When both $c_1$ and $c_2$ are $\hyperCoercionI$,
their composition is also $\hyperCoercionI$.  If the head of $c_2$ is
a projection, the result is $c_2$.  Otherwise, we need a label to
build the projection. Since the head of $c_2$ is $\epsilon$, its
source type must be a pretype. Thus we know $\ell$ must be a label and
we put it in the projection. When $c_1$ ends with a failure, the
composition is $c_1$ itself. In all the remaining cases, $c_1$ does
not end with a failure.  When $m_1$ and $m_2$ have different top
constructors ($m_1 \not\smile m_2$), the result is a failure coercion,
which needs a label. When $c_2$ starts with a projection, we blame the 
projection for casting a value to a shallowly inconsistent type. When $c_2$
starts with a $\epsilon$, we know $\ell$ must be a label, which is blamed for
composing shallowly inconsistent hypercoercions. The last two cases
compose $m_1$ and $m_2$ with the auxilliary function $m \fatsemi^\ell
m$, which assumes its inputs have the same top constructor. The
definition of $m \fatsemi^\ell m$ is a straightforward structural
recursion. Going back to the last two cases of $c_1 \fatsemi^\ell
c_2$. When $c_2$ starts with a projection, we do not know whether the
target type of $m_1$ matches the source of $m_2$. So we use the $l$ in the 
projection to  compose the middles $m_1 \fatsemi^l m_2$. When $c_2$ starts with 
a $\epsilon$, we compose the middles using $\ell$.

@@

A proof of correctness for hypercoercions (in 
Section~\ref{secc:framework:monoid-correct}) relies on the following
two basic properties of hypercoercions.

\begin{proposition}[\lazyD{} Hypercoercions Form a Monoid]
  \label{thm:hc-monoid}
  For all $c : T_1 \Longrightarrow T_2$,
  $c_1 : T_1 \Longrightarrow T_2$,
  $c_2 : T_2 \Longrightarrow T_3$, and
  $c_3 : T_3 \Longrightarrow T_4$,
  \begin{enumerate}
    \item $seq(id(T_1),c) = c$,
    \item $seq(c,id(T_2)) = c$, and
    \item $seq(seq(c_1, c_2), c_3) = seq(c_1, seq(c_2, c_3))$.
  \end{enumerate}
\end{proposition}
\begin{proof}
  Part (1) and (2) are by induction on $c$. Part (3) is by induction
  on $c_2$ and case analysis on $c_1$ then $c_3$.
\end{proof}

\begin{proposition}[\lazyD{} Identity Casts]
  For all $T$ and $l$, $cast(T,l,T) = id(T) $
\end{proposition}
\begin{proof}
  By induction on $T$.
\end{proof}

@@

\begin{proposition}[\lazyUD{} Hypercoercions Form a Monoid]
  For all $c : T_1 \Longrightarrow T_2$,
  $c_1 : T_1 \Longrightarrow T_2$,
  $c_2 : T_2 \Longrightarrow T_3$, and
  $c_3 : T_3 \Longrightarrow T_4$,
  \begin{enumerate}
    \item $seq(id(T_1),c) = c$,
    \item $seq(c,id(T_2)) = c$, and
    \item $seq(seq(c_1, c_2), c_3) = seq(c_1, seq(c_2, c_3))$.
  \end{enumerate}
\end{proposition}
\begin{proof}
  See the Agda proof at the following URL:

@@

\begin{proposition}[\lazyUD{} Identity Casts]
  For all $T$ and $l$, $cast(T,l,T) = id(T) $
\end{proposition}
\begin{proof}
  By induction on $T$.
\end{proof}

@@

\begin{proposition}[Equivalence of Two \lazyD{} Cast ADTs]
  \label{thm:surely-lazyD-eqv}
  Assume $C_1$ and $C_2$ implements Lazy D Cast ADT,
  \[
  eval_{\mathcal{S}(C_1)} = eval_{\mathcal{S}(C_2)}.
  \]
\end{proposition}
\begin{proof}
  Without lost of generality, assume $\judgeSreduce{C_1}{e}{o}$, prove 
  $\judgeSreduce{C_2}{e}{o}$.\\
  By the definition of $eval_{\mathcal{S}(C)}$, we know 
\[
\judgeSreduceTrans{C_2}{\sOOinspect{e}{\emptyset}{wrap(\kOOmt)}}{\sOOhalt{o}}
\]
  By induction on this reduction sequence and Lemma~\ref{thm:CEKS-bisim}, we 
  have 
\[
\judgeSreduceTrans{C_2}{\sOOinspect{e}{\emptyset}{wrap(\kOOmt)}}{\sOOhalt{o}}
\]
  that is, $\judgeSreduce{C_2}{e}{o}$
\end{proof}

@@

\begin{proposition}[$L$ Properties]\label{thm:L-LazyD-monoid}\ 
  \begin{enumerate}
  \item $L$ is an instance of the Lazy D ADT
  \item $L$ is a monoid
  \item $cast(T,l,T) = id(T)$
  \end{enumerate}
\end{proposition}
\begin{proof}
  Part (1) is straightforward because property (1) and (2) are trivial and 
  that $appS$ basically repeats property (3) - (11).
  Part (2) is true because in $L$ casts are list, $seq(c,c)$ is list 
  appending, and $id(T)$ is the empty list. Lists is know to be a monoid 
  where the binary operation is appending and the identity element is the 
  empty list.
  Part (3) follows immediately from the definition of $cast(T,l,T)$
\end{proof}

\begin{lemma}[$L$ is Correct wrt. \ineffCEKD]
  \label{lem:L-correct}
  \[
  \text{\evalEqv{\effCEK{L}}{\ineffCEKD}}
  \]
\end{lemma}
\begin{proof}
  Immedialy from Correctness of \effCEK{C} \ref{thm:surely-monoidic-eval} and 
  L properties \ref{thm:L-LazyD-monoid}.
\end{proof}

@@

\begin{theorem}[\effCEK{C} is Correct wrt. \ineffCEKD{}]
  \label{thm:surely-lazyD-correct}
  Suppose $C$ is a \lazyD{} Cast ADT.  If
  $\judgetype{\emptyset}{e}{T}$ and $o : T$, then
  \[
  \text{\evalEqv{\effCEK{C}}{\ineffCEKD}}
  \]
\end{theorem}
\begin{proof}
  If we have a Lazy D Cast ADT and it is a monoid, this 
  proof is immediately from Lemma~\ref{thm:surely-monoidic-eval} and 
  Proposition~\ref{thm:surely-lazyD-eqv}. One such representation is to 
  represent casts as a list of triples of type, label, and type, where 
  $id(T)$ is the empty list and $seq(c,c)$ is the list append. Another such 
  representation is the \lazyD\ hypercoercion!
\end{proof}

@@

\begin{lemma}[\lazyD{} Hypercoercion is a \lazyD{} Cast ADT]
  \label{thm:hc-surely-lazyD}
\end{lemma}
\begin{proof} See the supplementary material. \end{proof}

\begin{theorem}[\lazyD{} Hypercoercion Respect \ineffCEKD]
  If $\judgetype{\emptyset}{e}{T}$ and $o : T$
  \[
  \text{\evalEqv{\effCEK{H}}{\ineffCEKD}}
  \]
\end{theorem}
\begin{proof}
  Immediately from Theorem~\ref{thm:surely-lazyD-correct} and 
  Lemma~\ref{thm:hc-surely-lazyD}.
  Alternatively, from Lemma~\ref{thm:surely-monoidic-eval},
  Lemma~\ref{thm:hc-surely-lazyD}, and 
  Proposition~\ref{thm:hc-monoid}.
\end{proof}
